Diverse Genotype-by-Weather Interactions in Switchgrass

### 2024-07-17 manuscript/revision2/response_to_reviews/response_to_reviews.qmd

---
format:
  pdf:
    keep-tex: true
    pdfengine: xelatex
    geometry: 
      - "top=1in"
      - "bottom=1in"
      - "left=1in"
      - "right=1in"
    code-annotations: hover
---

::: {.flushright data-latex="1" linestretch="1"}
The University of Texas at Austin\
Department of Integrative Biology\
2415 Speedway\
Austin, TX 78712
:::

::: {.flushright data-latex="1" linestretch="1"}
July 17th, 2024
:::

::: {.flushleft data-latex="1" linestretch="1"}
Editors\
Proceedings of the National Academy of Sciences\

Subject: Manuscript revision
:::

Dear Editors,

@@

Thank you for considering this work for publication in *PNAS*. Our manuscript, "Diverse Genotype-by-Weather Interactions in Switchgrass" has undergone substantial revision from our original submission in 2021. This timeframe is long - Dr. MacQueen took a new position in 2022. They continued working with the group as we developed a new algorithm to address major reviewer concerns and significantly altered the manuscript. Because of the major alterations over this time frame, the response to reviewers does not have track changes for each point, though we do point to line numbers for the associated revisions in our response.

We were pleased by the useful feedback given by the reviewers of our original submission and believe that the greedy mash algorithm we implemented addresses the major statistical concern of reviewer 2, as well as substantially improving our use and interpretation of covariance matrices in *mash*. We have addressed all the editorial issues suggested by the two reviewers, and included in this letter are point-by-point responses to these issues.

We suggest that the original editor, Editorial Board members, and anonymous reviewers for this manuscript would be good choices to assess this revision. In the event that is not practicable, Douglas Schemske, Stephen P. Long, and Qifa Zhang would be good Editorial Board members for this manuscript. All would be able to provide strong insights into mapping the genetic basis of adaptive and ecologically relevant traits in a field experimental context. Johanna Schmitt, Trudy Mackay, and Detlef Weigel would be excellent editors for this manuscript. Dr. Schmitt has done extensive work in *Arabidopsis thaliana* and other plant species unravelling the genetic mechanisms involved in responses to the natural environment. Dr. Mackay has extensive experience mapping the genetics of complex traits and genotype-by-environment interactions. Dr. Weigel has done extensive work understanding genetic diversity in *A. thaliana* including extensive work on floral induction cues.

@@

> ::: {.callout-warning appearance="minimal"}
> Reviewer #1:\
> \
> Suitable Quality?: No\
> Sufficient General Interest?: No\
> Conclusions Justified?: Yes\
> Clearly Written?: No\
> Procedures Described?: Yes\
> \
> Comments on Significance Statement:\
> \
> Conclusions are very specific for the crop/population and traits. General conclusions are lacking.\
> \
> Comments:\
> \
> The paper by MacQueen et al. presents results of a study on two switchgrass populations grown in 8 garden experiments across a rane of latitudes. Traits are green-up date and flowering time. The data show strong GxE, effects of QTL changing in magnitude and sign. Results are partly validated in an independent segregating population.\
> \
> The topic is timely and highly relevant. A better understanding of GxE will be valuable for plant genetics, evolution and breeding. The data are complex and the authors have succeeded in making them accessible to the reader.\
> \
> \
> Relevant literature\
> The introduction and discussion are focused on findings in Arabidopsis and switchgrass. An excellent study from Arabidopsis with high relevance for this paper is\
> Fournier-Level et al. 2016 <https://doi.org/10.1073/pnas.1517456113>\
> \
> Major findings from other crop species have been ignored. To mention just two references:\
> \
> Bustos-Korts et al. 2019 The Plant Journal <https://doi.org/10.1111/tpj.14414>\
> Millet et al. 2016 Plant Physiology <https://doi.org/10.1104/pp.16.00621> and\
> There are many papers on crop modelling (eg work by Mark Cooper and others in maize) and the authors might want to consider linking their results to crop modelling to demonstrate the relevance of the results of this study
> :::

We now add these citations (lines 47-50; lines 58-61). We regret we can't comprehensively reference the impressive crop modeling literature. Our goal was to reference work on natural populations that focused the genetics on phenology and/or local adaptation, and to touch on how these ideas have interacted with crop improvement. This was because our work uses a natural population of switchgrass with potential relevance for breeding efforts in this species.

@@

We agree that we do not identify one general pattern or one environmental cue affecting any set of populations or phenological trait. Instead, our key expectation is that different populations and different genetic loci will have evolved distinct patterns of GxE (lines 55-58; lines 219-223; SI Appendix Section S1). Thus, we aim to identify the kinds of GxE present genome-wide and elaborate when these kinds of GxE differ (Figure 2), and demonstrate that individual loci have different types of GxE across environmental regions and subpopulations (Figure 3).

> ::: {.callout-warning appearance="minimal"}
> The authors claim that the environmental cues in the hypothesis based models improved model fit. What inference is possible from this result? Unless results are validated in independent data or with cross validation the predictive power of the cues cannot be judged.
> :::

We edit the Results to (i) change our analysis add an algorithm to more rigorously demonstrate this claim given the data that we have, and (ii) more clearly articulate this claim. We now employ a greedy mash algorithm to iteratively add covariance matrices to the *mash* model, only adding matrices that significantly improve the mash model log-likelihood, as we elaborate on further below in our [response to reviewer 2](#fig-greedy).

We do see ways we could ultimately use this for prediction - shrinking effects so that they have more accurate GxE would improve prediction, as in ([Zhu *et al.*, 2023](https://doi.org/10.1016/j.xgen.2023.100297)). However, prediction is beyond the scope of this manuscript; here we are introducing these ideas and using them to do two specific things: (i) find the GxE structures and (ii) look at the genome-wide prevalence of these structures. We feel that using mash to improve predictions is beyond the scope of this paper.

@@

> ::: {.callout-warning appearance="minimal"}
> Results from the SNPs "mash model of Midwest green-up fell on a covariance matrix of average temperature in the 10 days prior to Midwest green-up". These results should be linked to the phenotypic GxE analysis.
> :::

We edit the results to directly compare the GxWeather matrices and the phenotypic correlations in Figure 1B. (lines 241-258; lines 269-276; lines 283-290).

> ::: {.callout-warning appearance="minimal"}
> Flowering posterior weights were higher than green-up weights. What does that mean? How does that link to the different heritabilities and levels of GxE?
> :::

We believe this comment refers to there being higher loadings on the GxWeather matrices for flowering date than there were for the date of the start of vegetative growth in the original paper (original Figure 2). Our revised analyses [uses a greedy algorithm on canonical and GxWeather covariance matrices](#fig-greedy) to iteratively select only the subset that significantly improves the model log-likelihood. Thus, the covariance matrices included in each model have changed in this revision, as have the relative loadings on GxWeather and canonical covariance matrices.

We expected patterns of GxE and GxWeather to vary at the locus, subpopulation, and trait level, and we did not see a consistently higher fraction of green-up date mass than flowering date mass on the GxWeather matrices, nor vice versa. We don't believe it's appropriate to make direct comparisons between heritability or a variance components analysis and these mash loadings, so we do not do so in the paper; instead of the variance due to additive genetic variation and GxE, these show the additive genetic effects that most closely matched different patterns of GxE and GxWeather, regardless of the proportion of additive variance each locus has.

@@

No conclusions about individual loci effects can be drawn from the GxE analysis in Figure 2, as this plot does not show how shrinkage changes the jointly re-estimated SNP effect sizes. The GxE analysis in Figure 2 shows the types of GxE and GxWeather present genome-wide, from loadings on individual loci, but does not show the effects of these loci on traits. Rather, it is an overall characterization of GxE across all eight gardens.

The GxE analysis in Figure 3 shows the re-estimated effects of these loci on traits as contrasts between individual pairs of gardens. From Figure 3, you can conclude that many loci have effects with antagonistic pleiotropy at the site level.

> ::: {.callout-warning appearance="minimal"}
> Discussion\
> "we must understand the current patterns of trait covariation across environments, .....". What is the understanding from your data? Are there any general patterns or do we need to assess them individually for different genetic material, environments, traits? It looks like the latter so what are the consequences?
> :::

As we state in the SI Appendix (Section S1, paragraph two), we were interested in the types of genetic correlation that could be expressed by different alleles across the genome. We reasoned that different loci could have different patterns of genetic correlation. For example, the effects of an allele of locus $A$ could covary with a weather-based cue if $A$ had a common allele, $a$, that was responsive to this cue. In contrast, the effects of a different locus $B$ might have an effect at only one garden, for example if a pathogen was only present at one garden, and $B$ has a common allele $b$ that is resistant to that pathogen.

As a consequence, we develop an approach to specify multiple environmental cues and compete them to explain patterns of genetic effects (lines 230-237): we use a greedy algorithm to use with mash to select covariance patterns, and use mash to flexibly identify these patterns across populations and environments. (lines 238-240). With this approach, we are able to demonstrate that we can associate multiple patterns of GxWeather with specific genomic regions (lines 407-410). We can also assign genetic effects to both GxWeather patterns derived from weather variables and to other, site-based patterns (lines 410-412).

@@

> ::: {.callout-warning appearance="minimal"}
> "this is the first experimental work using QTL mapping and GWAS across ....". I disagree, see references above and others.
> :::

We edit the manuscript to remove this unclear claim. NB: We do cite the first experimental work using both QTL mapping and GWAS ([Brachi *et al.,* 2010](https://doi.org/10.1371/journal.pgen.1000940)), and there are many others; we meant, more specifically, using these approaches to map multiple types of GxE.

@@

> ::: {.callout-warning appearance="minimal"}
> "Gulf and Midwest subpopulations have two distinct photoperiod-related flowering responses....". This has been shown for other species, eg Unterseer et al. 2016 Genome Biology
> :::

We now reference this paper in the introduction (lines 47-52).

@@

> ::: {.callout-warning appearance="minimal"}
> Data/Methods\
> \
> Is one year data enough to make conclusions about the effect of environmental cues? It is well known that GxYear is more pronounced than GxLocation.
> :::

We agree that this would be an interesting topic for future work. Our aim was to demonstrate the value of this approach for mapping GxE, and we believe one year of data sufficient to do this.

> ::: {.callout-warning appearance="minimal"}
> The reader should be able to understand what the hypothesis in the hypothesis-driven models is without looking at the supplement.
> :::

Agreed. We add Table 1 to better explain the covariance matrices we select using the greedy mash algorithm. We include the matrices selected by this algorithm that have \>0.1% weight (lines 262-263) in the mash model as visualizations in Figure 2.

@@

Edited. We apologize for this omission.

> ::: {.callout-warning appearance="minimal"}
> Major issues:\
> \
> Conceptual issues:\
> \
> - I like the idea of using environmental measures near the time of phenological events to try to identify environmental drivers and compare the plasticity functions across environments. However the choices here do not make physiological sense, especially for flowering. The environmental indices focus on either the day of flowering, or the 1-2 weeks prior to flowering. However, the developmental transition to flowering likely occurs well before this interval (I'm not sure how long in switchgrass, but it's likely much earlier). Photoperiod and temperature cues driving this developmental transition are irrelevant once the developmental commitment has been made. The length of time before flowers emerge and open may be dependent on these environmental factors as well, but this is physiological, not related mechanistically to the daylength or temperature requirements for flowering. The intervals used may be more relevant to green-up, I'm not familiar with the developmental basis of this trait. But as is, I don't think these metrics are interpretable for what they are designed for.
> :::

We agree that the choice of environmental measure is an extremely important consideration, and we recognize that the time window the weather signal is integrated over could be long. We wanted our model to inform us on reasonable time windows, rather than having us assert these (probably incorrectly) in our model. Initially, we thought computational feasibility constrained the number of covariance matrices we could add to the mash model, as the runtime of mash increases with the number of covariance matrices included. To address this comment and the mash statistical issue the reviewer points out below, we now specify many more GxWeather covariance matrices with more time frames - 1-7 days prior to the phenological event, and 14, 21, and 28 days prior to the phenological event, generating 48-60 matrices per weather variable (Table 1). Then, we used a greedy mash algorithm, [explained below](#fig-greedy), to select GxWeather covariance matrices from this set that significantly improve the model likelihood. We thus extended the time frame for GxWeather covariance that our models could capture; however, only one 28 day matrix was selected by the greedy algorithm and had \>0.1% weight in our *mash* models.

We now also explicitly describe the two phenological traits more clearly in the text - both of the traits we map were measured on individuals, but specifically when approximately half of the ramets of the genotype had open flowers (for flowering) or green leaves (for green-up). Thus we are looking for cues driving vegetative and reproductive transitions for a majority of ramets, not for heading (flower emergence on the panicle). While there may be physiological mechanisms driving flower opening once the transition to reproductive development has been made, our results show that genetic variation underlies these differences in the timing of the physiological mechanisms underlying flower opening.

@@

> ::: {.callout-warning appearance="minimal"}
> \- GWAS: The GWAS model the authors use is not specified (nor was it specified in the 2021 Nature paper referenced). I believe looking at the source code that the model is a linear model with PCs to account for structure but no kinship matrix. It’s not clear how many PCs were used and since the raw GWAS results are not presented it’s impossible to tell if this was sufficient to account for the significant structure in these populations. I’m concerned about this because of the extremely unlikely number of significant markers (19K LD blocks). If each of these was a true positive, the average effect size of each block would be only 0.005% of the genetic variance. With only 350 accessions, the power to detect a locus that explains 10% of the genetic variance is only 50% with alpha = 10\^-5 and MAF = 0.05. So if the model is detecting tons of loci it’s likely that the model is severely biased by population structure, and these biases are likely to be somewhat consistent across gardens because the structure is similar. mash doesn’t help when the individual models are biased, it’ll find patterns of correlation whether they are biological or not and use those to shrink effect sizes together.
> :::

First, we now better document our GWAS methodology for this manuscript in Section S2 of the SI Appendix. We also include additional information on GWAS, including the number of PCs used and the Manhattan and QQ plots and associated data tables in the Github repository associated with the paper.

Second, though we selected 19K LD blocks in our set of 'strong' effects for *mash*, very few of these effects were significant (\<500 relatively unlinked LD blocks per set of eight gardens; this still represents some genomic inflation, no doubt, but not nearly as bad as 19K significant blocks would be). We now clarify this important detail in the main manuscript and in the SI Appendix.

Third, population structure could certainly bias the resultant mash models. We describe the steps we take to prevent "garbage in" to our mash models, including removing conditions where GWAS had significant population structure (SI Appendix, Section S2, last paragraph). However, it is our intuition that, should biases remain in our models, as they undoubtedly do, mash would not load these biased effects onto our GxWeather covariance matrices - instead, these effects likely load onto canonical matrices, such as garden-specific effects, or potentially onto data-driven matrices. Thus, we expect our assessment of the proportion of the genome that shows GxWeather effects to be conservative if our models include residual population structure.

@@

> ::: {#comment-cov-correlations .callout-warning appearance="minimal"}
> \- Mash: A lot of the results interpretation is based on interpreting the SNP loadings on the specified covariance matrices. This is a secondary use of mash (the primary being the refinement of effect sizes), and while they do interpret these somewhat in the mash paper, these loadings need to be interpreted with caution. If these covariance matrices are similar, mash will somewhat arbitrarily assign weight to any of the matrices because all lead to the same fit to the effect-size data. The mash algorithm doesn’t give the full posterior distribution on the loadings so you can’t check for posterior correlations there. It seems likely to me that the hypothesis-driven and data-driven covariance matrices are somewhat correlated here, and the correlations may differ between green-up and flowering because of the better correlation between the environmental metrics and flowering than green-up. In the mash paper, they used cross-validation with the likelihood in the testing set as the evaluation metric to compare models. It might be safer to try dropping specific covariance matrices and comparing the model performance in a held-out testing set to evaluate the importance of the hypothesis-derived covariance matrices.
> :::

We thank the reviewer for this comment and agree that the selection of covariance matrices to be included in a mash model is a nontrivial question. To clarify our interpretation of the posterior matrix weights provided by mash, we performed an extensive analysis of the performance of mash models when different covariance matrices are included. Specifically, we implemented a model selection approach that uses a greedy algorithm to evaluate the log likelihood of the mash model as additional covariance matrices were included (see Methods lines 899-906, SI Appendix Section S4, and pseudo code below). This is a similar, but more computationally efficient design, than the leave-one-out approach recommended by the reviewer as the runtime of mash increases with the number of covariance matrices included. Additionally, while cross-validation is a powerful approach to model evaluation our limited sample size was prohibitive to the necessary partitioning of individuals included in our analysis.

In practice this greedy algorithm approach offers a fast way to identify the point where redundancy in the addition of a new covariance matrix results in no change to the likelihood of the mash model. Importantly, this removes the potential for arbitrary assignment of weights to the covariance matrices in the model and allows for the selection of matrices that most accurately capture the underlying biology captured by the site-specific effect sizes. We applied this algorithm to only the GxWeather covariance matrices for each phenotype (Figures 1 and 2) using the previously used sets of significantly associated and randomly selected variants in the Gulf subpopulation individuals, Midwest subpopulation individuals, and the two subpopulations combined. The results indicate that only a small number of the original hypothesis matrices (between three and five) are necessary to reach the maximum likelihood model when using the same likelihood ratio test implemented in Urbut *et al*., indicating that correlation among hypothesis covariance matrices is high. This redundancy was confirmed through our application of the greedy algorithm to a combined set of canonical, data driven, and GxWeather covariance matrices (Figures 3 and 4) Our primary conclusion from this analysis is in line with what the reviewer posited, that there is extensive redundancy among canonical, data driven, and hypothesis covariance matrices that is leading to biologically informative patterns of covariance between site effect sizes to be arbitrarily distributed among them. We additionally found evidence that while the correlations between data driven and environmentally informed (hypothesis) covariance matrices, there is extensive similarity in the matrices that are included in the maximum likelihood models across both phenotypes and in the analysis of the Gulf subpopulation, Midwest subpopulations, and the combined cohort.

@@

  matrices.remove(most_likely_matrix)
  most_likely_matrices.append(most_likely_matrix)
  maximum_likelihoods.append(likelihood)

  lrt_pvalue = likelihood_ratio_test(likelihood, maximum_likelihoods[iteration-1], df = 1)
____________________________________________________________________________
```

### 2024-07-17 manuscript/revision2/SI_Appendix/SI_Appendix.qmd
revision2: Sync cross-references between revision, SI Appendix, and response letter.
Also, add Figures to response letter and change templating to allow this

To create the GxWeather covariance matrices we transformed each genotype's green-up and flowering date at each garden to create date-related values that were summary functions for a weather variable across a specified date range. For example, the cumulative sum of the rainfall that occurred seven days prior to the flowering date was used to create one possible GxWeather cue variable for flowering date. The code for these calculations can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 394.

@@

We then generated GxWeather covariance matrices derived from correlations in these GxWeather cue variables for the three subpopulations. These covariance matrices represent the correlations between genotypes for these weather-related variables across our eight common gardens. To create these matrices, we determined the correlations between these weather-related cue variables for identical genotypes grown at different common gardens and used these correlations to fill the off-diagonals of a eight-by-eight matrix. To fill the diagonal entries of these matrices, we used the narrow-sense heritability for each weather-related cue variable at each garden, calculated using the same methods as Section S3. Matrices need to be positive semi-definite to be used in *mash*, so when the resultant matrices were not positive semi-definite, the `Matrix::nearPD` function was used to compute the nearest positive definite matrix to the correlation matrix [@Matrix]. Finally, all covariance matrices were rescaled so that the maximum diagonal element of the rescaled matrix is 1, as in [@Urbut2019]. Code used to generate these matrices can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe.git/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 720.

@@

where the vector $\mathbf{y}$ represents the flowering date or greenup date values for that garden, $Z$ the design matrix for random effects, $u$ the whole genome additive genetic effect, $e$ the residual, $G$ the genomic relationship matrix based on all SNPs retained for subpopulation-specific analyses, and $I$ the rank-$\mathbf{y}$ identity matrix [@clark2013]. ASReml can specify considerably more complex variance structures; however, we were interested in specifying covariance structures using *mash*, which allows the specification of multiple covariance structures; hence, we used a relatively simple model to calculate BLUPs for use in univariate GWAS. The code for these calculations can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe.git/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 75.

Univariate GWAS were run independently for each combination of trait, genetic subpopulation, and common garden. The code used to run GWAS can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe.git/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 240. We used [`pvdiv_standard_run()`](https://alice-macqueen.github.io/switchgrassGWAS/reference/pvdiv_standard_gwas.html), an opinionated function of the [switchgrassGWAS](https://github.com/Alice-MacQueen/switchgrassGWAS) R package, to conduct GWAS. To control for population structure and run GWAS, this function uses a three-part process. First, a fast truncated singular value decomposition (SVD) is generated from the SNP data, which has initial pruning of regions in long-range linkage disequilibrium (LD) and iterative pruning and removal of regions in long-range LD [@privé2017a]. By default, 15 vectors of singular values, $k$, are computed. Pruning and removal of long-range LD regions are recommended in research in human genetics to best control for population structure [@abdellaoui2013; @privé2017; @price2008]. Second, linear regression models using 0-15 of these vectors of singular values were used to run GWAS and to calculate the genomic control coefficient, $\lambda_{GC}$, for each $k$ [@Devlin1999]. Third, we chose the lowest value of $k$ that made $\lambda_{GC}$ closest to 1, and ran GWAS using a linear regression model with this $k$ for population structure correction.

At this stage, we reviewed the univariate GWAS results, both Manhattan and QQ-plots, for population structure issues that were not resolved using the $k$ that minimized $\lambda_{GC}$. These issues were apparent in the data in two ways: (i) univariate GWAS with many significant SNPs with extremely similar *p*-values, usually indicative of genetic effects that, by chance or, more likely, because of confounding effects of population structure, were identical across many SNPs and thus not resolvable to a single genomic location; (ii) univariate GWAS with many effect estimates and standard errors of zero, indicating an inability to estimate effects given the constraints of small population size & population structure. We only noticed these issues in our smallest subpopulation, the Midwest subpopulation, in the green-up date trait at the SD garden, and in the flowering date trait at the OK and NE gardens. Due to these issues, we dropped the SD garden from subsequent green-up analyses for the Midwest subpopulation, and we dropped the OK and NE gardens from subsequent Midwest flowering date analyses.

@@

@@

The code used to run *mash* can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe.git/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 1326.

>>>>

To create the GxWeather covariance matrices we transformed each genotype's green-up and flowering date at each garden to create date-related values that were summary functions for a weather variable across a specified date range. For example, the cumulative sum of the rainfall that occurred seven days prior to the flowering date was used to create one possible GxWeather cue variable for flowering date. The code for these calculations can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 387.

@@

We then generated GxWeather covariance matrices derived from correlations in these GxWeather cue variables for the three subpopulations. These covariance matrices represent the correlations between genotypes for these weather-related variables across our eight common gardens. To create these matrices, we determined the correlations between these weather-related cue variables for identical genotypes grown at different common gardens and used these correlations to fill the off-diagonals of a eight-by-eight matrix. To fill the diagonal entries of these matrices, we used the narrow-sense heritability for each weather-related cue variable at each garden, calculated using the same methods as Section S3. Matrices need to be positive semi-definite to be used in *mash*, so when the resultant matrices were not positive semi-definite, the `Matrix::nearPD` function was used to compute the nearest positive definite matrix to the correlation matrix [@Matrix]. Finally, all covariance matrices were rescaled so that the maximum diagonal element of the rescaled matrix is 1, as in [@Urbut2019]. Code used to generate these matrices can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe.git/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 715.

@@

where the vector $\mathbf{y}$ represents the flowering date or greenup date values for that garden, $Z$ the design matrix for random effects, $u$ the whole genome additive genetic effect, $e$ the residual, $G$ the genomic relationship matrix based on all SNPs retained for subpopulation-specific analyses, and $I$ the rank-$\mathbf{y}$ identity matrix [@clark2013]. ASReml can specify considerably more complex variance structures; however, we were interested in specifying covariance structures using *mash*, which allows the specification of multiple covariance structures; hence, we used a relatively simple model to calculate BLUPs for use in univariate GWAS. The code for these calculations can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe.git/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 69.

Univariate GWAS were run independently for each combination of trait, genetic subpopulation, and common garden. The code used to run GWAS can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe.git/R/Analysis_v1.2_mash_using\_ greedy_covar.qmd, starting at line 240. We used [`pvdiv_standard_run()`](#0), an opinionated function of the [switchgrassGWAS](#0) R package, to conduct GWAS. To control for population structure and run GWAS, this function uses a three-part process. First, a fast truncated singular value decomposition (SVD) is generated from the SNP data, which has initial pruning of regions in long-range linkage disequilibrium (LD) and iterative pruning and removal of regions in long-range LD [@privé2018]. By default, 15 vectors of singular values, $k$, are computed. Pruning and removal of long-range LD regions are recommended in research in human genetics to best control for population structure [@abdellaoui2013; @privé2018; @price2008]. Second, linear regression models using 0-15 of these vectors of singular values were used to run GWAS and to calculate the genomic control coefficient, $\lambda_{GC}$, for each $k$ [@Devlin1999]. Third, we chose the lowest value of $k$ that made $\lambda_{GC}$ closest to 1, and ran GWAS using a linear regression model with this $k$ for population structure correction.

At this stage, we reviewed the univariate GWAS results, both Manhattan and QQ-plots, for population structure issues that were not resolved using the $k$ that minimized $\lambda_{GC}$. These issues were apparent in the data in two ways: (i) univariate GWAS with many significant SNPs with extremely similar *p*-values, usually indicative of genetic effects that, by chance or, more likely, because of confounding effects of population structure, were identical across many SNPs and thus not resolvable to a single genomic location; (ii) univariate GWAS with many effect estimates and standard errors of zero, indicating an inability to estimate effects given the constraints of small population size & population structure. We only noticed these issues in our smallest subpopulation, the Midwest subpopulation, in the green-up date trait at the MO and SD garden, and in the flowering date trait at the OK and NE gardens. Due to these issues, we dropped the MO and SD garden from subsequent green-up analyses for the Midwest subpopulation, and we dropped the OK and NE gardens from subsequent Midwest flowering date analyses.

@@

The intuition behind taking a subset of relatively unlinked, randomly selected effects is that this will capture a realistic fraction of the null and non-null effects in the full dataset, and thus be an unbiased representation of the amount and types of signal in the full dataset. Mash then uses this subset of random effects to estimate the number of null effects in the dataset and shrink some estimates towards null effects, effectively, towards a covariance matrix with no effects. The intuition behind using a subset of relatively unlinked effects with strong effects, or low *p*-values, is that these effects are the most likely to end up with significant jointly re-estimated SNP effects, and every SNP in a linkage block is likely to have effects re-estimated similarly.

@@

The code used to run *mash* can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe.git/R/Analysis_v1.2_mash\_ using_greedy_covar.qmd, starting at line 1340.
